\documentclass[12pt]{book}
\usepackage{style}

\title{Spectral Graph Theory}
\author{Lorenzo Orecchia, Erasmo Tani}

\begin{document}


\iffalse
This is one traditional page order for books.

Frontmatter

   - Half-title
   - Empty
   - Title page
   - Information (copyright notice, ISBN, etc.)
   - Dedication if any, else empty
   - Table of contents
   - List of figures (can be in the backmatter too)
   - Preface chapter

Mainmatter

   - Main topic

Appendix

    - Some subordinate chapters

Backmatter

    - Bibliography
    - Glossary / Index


 Chapters after the frontmatter are not numbered
 Page numbers will be printed in roman numerals.
 Frontmatter is not supposed to have sections, so they will be numbered 0.n because there is no chapter numbering.

\fi

\frontmatter

\maketitle

% Introductory chapters
\chapter{Preface}

% ----------------------------------------- %
% ------------ End of Preface  ------------ %
% ----------------------------------------- %


\iffalse

  The mainmatter chapters works as usual. The command resets the page numbering.
  Page numbers will be printed in arabic numerals.

\fi

\mainmatter
\chapter{First chapter}

\include{tex/RandomWalks}

% when \include is used, the extension should be avoided, otherwise the project won't compile the additional file.
\include{tex/PSD}



% ------------------------------------------ %
% ---------- End of Main Content  ---------- %
% ------------------------------------------ %

\iffalse

   The \appendix macro can be used to indicate that following
   sections or chapters are to be numbered as appendices.
   Appendices can be used for the article class too:

\fi

\appendix
\chapter{First Appendix}

\iffalse

   The backmatter behaves like the frontmatter.
   It has the same issue with section numbering.

\fi

\backmatter
\chapter{Last note}



\end{document}
